\Title

### 2013-06-05 synmut.tex
little changes

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
In HIV-1, because of the high mutation rate, single nucleotide variants (SNVs)
are continuously injected into the viral population. We say that an SNV has a
frequency $\nu$ at a certain time if it is found in a fraction $\nu$, between
zero and one, of the sequences from that time point. An SNV rises or falls in
frequency because of (i) its own effect on fitness and escape and (ii)
its association to genetic backgrounds and (iii) stochastic fluctuations
(genetic drift). When an SNVs is present in all sequences
at a certain time point, we say it has ``fixed''; when it is completely absent,
we say it was ``lost'' or is ``extinct''.

Eventually, all SNVs will either fix or be lost, at least in small
samples. Given the SNV is at a certain frequency $\nu$, the probability
of fixation is higher for beneficial SNVs than for neutral ones, which
in turn fix more frequently than deleterious ones. Since a neutral
mutation, by definition, has no effect on the fate of the viruses
carrying it, the fixation probability of a neutral SNV at frequency
$\nu$ is $\pfix(\nu) = \nu$; it goes extinct with probability
$1-\nu$. For instance, if a neutral SNV is observed in half of the
sequences, it will fix with a probability of 50\% (see inset in figure
\ref{fig:aftsyn}). The fixation probability of neutral SNVs is
independent of most model assumptions and is only affected if neutral
SNVs are associated preferentially with either high or low fitness
virus.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Many synonymous SNVs in \shankaregion{} are deleterious}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We studied the dynamics and fate of synonymous SNVs more quantitatively by
analyzing data from seven patients from \citet{shankarappa_consistent_1999} and
\citep{liu_selection_2006} as well as three patients from
\citet{bunnik_autologous_2008} (patients with strong viral population structure
were excluded from the analysis; see methods and \figurename~S\PCApat). The
former data set is restricted to the \shankaregion{} region of \env, while the
data from \citet{bunnik_autologous_2008} cover most of \env.  We considered all
SNVs in a frequency interval $[\nu_0-\delta\nu, \nu_0+\delta\nu]$ at some time
$t$, and calculated the fraction that are still observed at later times $t+\Delta
t$. Plotting this fraction against the time interval $\Delta t$, we see that
most synonymous SNVs segregate for roughly one year and are lost much more
frequently than expected under neutrality (panel \ref{fig:fixp1}). The long-time
probability of fixation, $\pfix$, is shown as a function of the initial
frequency $\nu_0$ in panel \ref{fig:fixp2}. We found that $\pfix$ of synonymous
variants is far below the neutral expectation in \shankaregion (red line).
Outside of \shankaregion, using data from \citet{bunnik_autologous_2008} only,
we found no such reduction in $\pfix$ (green line). Restricted to the
\shankaregion{} region, the sequence samples from \citet{bunnik_autologous_2008}
are fully compatible with data from \citet{shankarappa_consistent_1999}. The
nonsynonymous SNVs seem to follow more or less the neutral expectation (blue
line) -- a point to which we will come back below.

@@

The above simulations show that hitchhiking can explain the observation of
deleterious SNVs that rarely fix. However, in a simple model where
nonsynonymous escape mutations are unconditionally beneficial, they almost
always fix once they reach high frequencies, i.e., $A_{\mathrm{nonsyn}}$ is well
above zero. This is incompatible with the blue line in \FIG{fixp2}: in an HIV-1
infection, nonsynonymous SNVs at high frequency often disappear again, even
though many are at least transiently beneficial. Inspecting the trajectories of
nonsynonymous SNVs suggests the rapid rise and fall of many SNVs. Two
possible mechanisms that could explain the transient rise of
nonsynonymous SNVs are time-dependent selection and within-epitope competition.

If the immune system
recognizes the escape mutant before its fixation, the mutant might cease to be
beneficial and disappear soon, despite its quick initial rise in frequency. In
support of this idea, in \citet{richman_rapid_2003, bunnik_autologous_2008},
antibody responses to escape mutants have been reported. These responses are
delayed by a few months, roughly matching the average time needed by an escape
mutant to rise from low to high frequency. To model this type of behavior, we
assumed that antibody responses against escape SNVs arise with a rate
proportional to the frequency of the escape SNV and abolish the benefit of the
escape mutations. As expected, this type of time-dependent selection retained the
potential for hitchhiking, but reduced fixation of nonsynonymous SNVs.
\figurename~S\timedependence~shows that $\pfix$ of synonymous SNVs is not
affected by this change, while $\pfix$ of nonsynonymous SNVs approaches the
diagonal as the rate of recognition of escape mutants is increased.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
By analyzing the fate of single nucleotide variants (SNVs) in
longitudinal data of HIV-1 \env{} evolution, we demonstrated selection
against synonymous substitutions in the relatively conserved regions
C2-C4 of the \env{} gene. Computational modeling suggested that these
SNVs have deleterious effects on the order of $0.002$ and that they are
brought to high frequency through linkage to adaptive mutations.
Comparison with biochemical studies of base pairing propensity in RNA
genome of HIV-1 indicated that these mutations are deleterious, at least
in part, because they disrupt RNA secondary structures. Computer models
of RNA folding predict stable hairpins in these regions that
have been suggested to be functional and termed ``insulating
stems'' \citep{watts_architecture_2009, sanjuan_interplay_2011}.
The weak selection against synonynous variants is compatible with the
negative results of \textit{In vitro} replication assays investigating
the fitness effects of small RNA hairpins in HIV-1
\citep{knoepfel_role_2013}: it would take hundreds of cell culture
passages to detect fitness effects of the order of one per mille. The longitudinal data,
however, spans many years and our analysis is able to quantify the
subtle fitness effect of RNA structure within single infections.

The fixation probabilities and the sojourn times of SNVs represent a rich and simple
summary statistics useful to characterize longitudinal sequence data and compare
to models via computer simulations. These statistics are informative
even in absence of a neutral control and thus suited to analyze
properties of synonymous sites. 
We find selection against synonymous substitutions  despite the
fact that the corresponding sites are not conservered in
cross-sectional data. This is consistent with a recent comparative
analysis of SIV and HIV-1 RNA secondary structure using SHAPE assays and
computational methods \citep{pollom_comparison_2013}. While large scale
patterns of RNA structures tend to agree in both viruses, the individual
base pairs forming the structures are almost always
discordant. Even though the molecular architecture of the structure
changes over time, selection seems to maintain these structures, which reduces the
fixation probability and hence the rate of evolution at synonymous
sites. As expected from this argument, the evolutionary rate
at synonymous sites varies greatly along the HIV-1 genome
\citep{mayrose_towards_2007}. This variation can confound estimates of
selection on proteins substantially \citep{ngandu_extensive_2008}. 
The dynamic nature of HIV secondary structure makes finding the 
compensatory mutations that would restore base pairing in the
longitundinal data difficult, since the exact base pairing pattern is
most likely different than in the reference sequence.

@@

The SHAPE scores quantifying the degree of base pairing of individuals sites in
the HIV-1 genome were downloaded from the journal website
\citep{watts_architecture_2009}. Wherever possible, SHAPE reactivities were
assigned to sites in the multiple sequence alignments for each patient through
the alignment to the sequence of the NL4.3 virus used in
\citep{watts_architecture_2009}. Problematic assignments in indel-rich
regions were excluded from the analysis. The variable loops and flanking
regions were identified manually starting from the annotated reference HXB2
sequence from the LANL HIV database~\citep{LANL2012}.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Due to the large population size and the high mutation rate, every
possible single nucleotide variant (SNV) is produced multiple time per
day \citep{coffin_hiv_1995}. Some of these SNVs rise to high enough
frequency that they are observed in a sample of sequences. SNVs rise or fall in 
frequency because of (i) their own effect on fitness and escape and (ii)
their association to genetic backgrounds and (iii) stochastic fluctuations
(genetic drift). We study the dynamics of sample frequencies, $\nu$, of SNVs
(SNV), i.e., the fraction $\nu$ of the sequences in a sample carrying
the SNV. When an SNVs is present in all sequences 
at a certain time point, we say it has ``fixed''; when it is completely absent,
we say it was ``lost'' or is ``extinct''.

Most positions are only transiently variable and SNVs will either fix or
are lost -- at least in small samples. Given a SNV is at a certain
frequency $\nu$, the probability of fixation is higher for beneficial
SNVs than for neutral ones, which in turn fix more frequently than
deleterious ones. Since a neutral mutation, by definition, has no effect
on the fate of the viruses carrying it, the fixation probability of a
neutral SNV at frequency $\nu$ is the frequency itself, i.e.,
$\pfix(\nu) = \nu$; it goes extinct with probability $1-\nu$. For
instance, if a neutral SNV is observed in half of the sequences, it will
fix with a probability of 50\% (see inset in figure
\ref{fig:aftsyn}). The fixation probability of neutral SNVs is
independent of most model assumptions and is only affected if neutral
SNVs are associated preferentially with either high or low fitness
virus.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Many synonymous SNVs in \shankaregion{} are deleterious}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We studied the dynamics and fate of synonymous SNVs more quantitatively by
analyzing data from seven patients from \citet{shankarappa_consistent_1999} and
\citet{liu_selection_2006} as well as three patients from
\citet{bunnik_autologous_2008} (patients with strong viral population structure
were excluded from the analysis; see methods and \figurename~S\PCApat). The
former data set is restricted to the \shankaregion{} region of \env, while the
data from \citet{bunnik_autologous_2008} cover most of \env.  We considered all
SNVs in a frequency interval $[\nu_0-\delta\nu, \nu_0+\delta\nu]$ at some time
$t$, and calculated the fraction that are still observed at later times $t+\Delta
t$. Plotting this fraction against the time interval $\Delta t$, we see that
most synonymous SNVs segregate for roughly one year and are lost much more
frequently than expected under neutrality (panel \ref{fig:fixp1}). The long-time
probability of fixation, $\pfix$, is shown as a function of the initial
frequency $\nu_0$ in panel \ref{fig:fixp2}. We found that $\pfix$ of synonymous
variants is far below the neutral expectation in \shankaregion (red line).
Outside of \shankaregion, using data from \citet{bunnik_autologous_2008} only,
we found no such reduction in $\pfix$ (green line). Restricted to the
\shankaregion{} region, the sequence samples from \citet{bunnik_autologous_2008}
are fully compatible with data from \citet{shankarappa_consistent_1999}. The
nonsynonymous SNVs seem to follow more or less the neutral expectation (blue
line) -- a point to which we will come back below.

@@

The above simulations show that hitchhiking with favored nonsynonymous
variants can explain the observation of
deleterious synonymous SNVs that rarely fix. However, in a simple model where
nonsynonymous escape mutations are unconditionally beneficial, they almost
always fix once they reach high frequencies, i.e., $A_{\mathrm{nonsyn}}$ is well
above zero. This is incompatible with the blue line in \FIG{fixp2}: in an HIV-1
infection, nonsynonymous SNVs at high frequency often disappear again, even
though many are at least transiently beneficial. Inspecting the trajectories of
nonsynonymous SNVs suggests the rapid rise and fall of many SNVs. Two
possible mechanisms that could explain the transient rise of
nonsynonymous SNVs are time-dependent selection and within-epitope competition.

If the immune system
recognizes the escape mutant before its fixation, the mutant might cease to be
beneficial and disappear soon, despite its quick initial rise in frequency. In
support of this idea, in \citet{richman_rapid_2003, bunnik_autologous_2008},
antibody responses against escape mutants have been reported. These responses are
delayed by a few months, roughly matching the average time needed by an escape
mutant to rise from low to high frequency. To model this type of behavior, we
assumed that antibody responses against escape SNVs arise with a rate
proportional to the frequency of the escape SNV and abolish the benefit of the
escape mutations. As expected, this type of time-dependent selection retained the
potential for hitchhiking, but reduced fixation of nonsynonymous SNVs.
\figurename~S\timedependence~shows that $\pfix$ of synonymous SNVs is not
affected by this change, while $\pfix$ of nonsynonymous SNVs approaches the
diagonal as the rate of recognition of escape mutants is increased.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
By analyzing the fate of single nucleotide variants (SNVs) in
longitudinal data of HIV-1 \env{} evolution, we demonstrated selection
against synonymous substitutions in the relatively conserved regions
C2-C4 of the \env{} gene. Computational modeling suggested that these
SNVs have deleterious effects on the order of $0.002$ and that they are
brought to high frequency through linkage to adaptive mutations.
Comparison with biochemical studies of base pairing propensity in RNA
genome of HIV-1 indicated that these mutations are deleterious, at least
in part, because they disrupt RNA secondary structures. Computer models
of RNA folding predict stable hairpins in these regions that
have been suggested to be functional and termed ``insulating
stems'' \citep{watts_architecture_2009, sanjuan_interplay_2011}.
The weak selection against synonymous variants is compatible with the
negative results of \textit{In vitro} replication assays investigating
the fitness effects of small RNA hairpins in HIV-1
\citep{knoepfel_role_2013}: it would take hundreds of cell culture
passages to detect fitness effects of the order of one per mille. The longitudinal data,
however, spans many years and our analysis is able to quantify the
subtle fitness effect of RNA structure within single infections.

The fixation probabilities and the sojourn times of SNVs represent a rich and simple
summary statistics useful to characterize longitudinal sequence data and compare
to models via computer simulations. These statistics are informative
even in absence of a neutral control and thus suited to analyze
properties of synonymous sites. 
We find selection against synonymous substitutions  despite the
fact that the corresponding sites are not conserved in
cross-sectional data. This is consistent with a recent comparative
analysis of SIV and HIV-1 RNA secondary structure using SHAPE assays and
computational methods \citep{pollom_comparison_2013}. While large scale
patterns of RNA structures tend to agree in both viruses, the individual
base pairs forming the structures are almost always
discordant. Even though the molecular architecture of the structure
changes over time, selection seems to maintain these structures, which reduces the
fixation probability and hence the rate of evolution at synonymous
sites. As expected from this argument, the evolutionary rate
at synonymous sites varies greatly along the HIV-1 genome
\citep{mayrose_towards_2007}. This variation can confound estimates of
selection on proteins substantially \citep{ngandu_extensive_2008}. 
The dynamic nature of HIV secondary structure makes finding the 
compensatory mutations that would restore base pairing in the
longitundinal data difficult, since the exact base pairing pattern is
most likely different than in the reference sequence.

@@

The SHAPE scores quantifying the degree of base pairing of individuals sites in
the HIV-1 genome were downloaded from the journal website
\citep{watts_architecture_2009}. Wherever possible, SHAPE reactivities were
assigned to sites in the multiple sequence alignments for each patient through
the alignment to the sequence of the NL4-3 virus used in
\citep{watts_architecture_2009}. Problematic assignments in indel-rich
regions were excluded from the analysis. The variable loops and flanking
regions were identified manually starting from the annotated reference HXB2
sequence from the LANL HIV database~\citep{LANL2012}.
